---
title: "Project Report Statistics"
subtitle: "Protests and Campaigns"
author: "Christian Vorhauser"
#format: html-notebook
#editor: visual
title-block-banner: "#42504d"
title-block-banner-color: "white"
format: 
  html-notebook:
    embed-resources: true
    smooth-scroll: true
    theme: cosmo
    fontcolor: black
    toc: true
    toc-location: left
    toc-title: Table of Contents
    toc-depth: 3
css: style.css
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(car)
library(tidyverse)
library(countrycode)
library(gamlr)
```

# Describing the data

The dataset I chose for my project includes 299 campaigns/projects from all around the world in the time frame of 1960 to 2013. The response variable is the success of the campaign, originally encoded as TRUE or FALSE. But when was a campaign successful? A protest (most often) has a clear objective behind it, be it the dismantlement of the reigning government, or convincing them to not put a law in practice. These are also the only kinds of campaigns present in the data set, and thus a success is granted by the achievement of this goal.

The variables available for each of the campaigns include the participation of the population in percent and the absolute number of participants during the height of it. For some campaigns we have participation rates of less than 0.01% of the population all the way up 128%. This can be the case when people from a different, mainly neighboring, country take part in the campaign and are thus counted towards this number. Since the campaign is only concerning a single country it is possible to achieve values of this magnitude.

The next column is an identifier, which bears no information for us and we thus disregard it. 

Following this, three columns regarding the time of the campaign are listed. The beginning year, the ending year and the year when it was at it's height, meaning the highest number of participants were recorded. As mentioned before we have years ranging from 1960 to 2013, with many taking place either in the time frame 1987-1992, or between 2009-2013. The earlier period can most likely be associated with the fall of the USSR around 1991 where many countries gained their independence. I assume a higher number of campaigns during that time, as the people protested against certain parties, which would have taken control, or be part of the fall itself by campaigning against the USSR government.
The second period could show this increase due to the "Arab Spring" in the early 2010s where many countries in the middle east and northern Africa revolted against the (authoritarian) government. This can also explain the end of many campaigns in 2013 as the "official" end of it was December 2012 (according to Wikipedia). 

To support the theories postulated in the above paragraph, the next column gives us the location of the campaign, which reveals to us that, indeed these two events caused a spike in campaigns during the mentioned time periods. Since quite a number of countries split or got created in the 50+ years our data spans, we need to make sure that, when analyzing it later we treat them correctly regarding continent they are on, since some libraries will not have former countries in their database.

Coming back to a population metric, the next variable is the population in the concerned country during the height of the campaign. This is also the number of citizens used when calculating the participation in terms of population mentioned beforehand. 

Lastly, we columns concerned with, on one hand, if more than 3.5% of the population were engaged in the campaign, and on the other hand, whether the protest was peaceful, or violent. The last one is encoded as "chill" and "PROBLEM" in the original data. For a protest to violent there have to have been physically injured people, it does not matter whether they are on the side of the participants, government/police/military, or the civilian population. 

Coming to the research question I will try to answer: Is violence a significant component to the success of a campaign, or are other factors more influential? Other factors present in the data set that could have an influence are the number of people participating, the share of the population they make up, the duration of the campaign or the location of it.

# Data Processing and Transformations

To make the data suitable to work with there were some changes to be made. First of all the categorical variables need to be changed to 0/1 encoding. For the singular variables this means:
- A successful campaign is 1, unsuccessful 0
- More than 3.5% involved is 1, less is 0
- A violent campaign is 1, a peaceful one 0

The next step is to change the data type of the variables to numeric when needed. For the participation in percent this means removing the percentage sign ("%") from the string and then converting to numeric. For the population numbers the "," had to be removed, since R uses "." for decimal and nothing for separation of thousands. Afterwards, the conversion to numerics was possible.

Since, I had more than 100 distinct countries in the location column and creating this many dummy variables is not feasible or helps us explain our research question, I decided to assign them each to their respective continent.
This was done with the "countrycodes" package. There were five locations that could not be assigned a continent unambiguously, so I investigated further and found that they were Czechoslovakia and three different versions of Yemen. Czechoslovakia was assigned Europe, and all the Yemens got Asia. 

```{r}
#| echo: false
#| message: false
#| include: false


# laoding and preparing the data
data <- read.csv("protests.csv",header = TRUE,skip = 1)
# chaniging treu false to 0,1
data[which(data$Success. == FALSE),"Success."] <- 0
data[which(data$Success. == TRUE),"Success."] <- 1

data[which(data$greater.than.3.5..of.pop.involved. == FALSE),"greater.than.3.5..of.pop.involved."] <- 0
data[which(data$greater.than.3.5..of.pop.involved. == TRUE),"greater.than.3.5..of.pop.involved."] <- 1

data[which(data$Problem. == "chill"),"Problem."] <- 0
data[which(data$Problem. == "PROBLEM"),"Problem."] <- 1

# bringing the variables into the right format (mostly datatype)
data$participants.as...population <- gsub("%","",data$participants.as...population)
data$participants.as...population <- as.double(data$participants.as...population)

data$MAX.of.participants <- gsub(",","",data$MAX.of.participants)
data$MAX.of.participants <- as.numeric(data$MAX.of.participants)

data$Pop.in.that.year <- gsub(",","",data$Pop.in.that.year)
data$Pop.in.that.year <- as.numeric(data$Pop.in.that.year)

data$Problem. <- as.numeric(data$Problem.)

# grouping countrie into continents

data$continent <- countrycode(data$Location,"country.name","continent")
data$continent[which(data$Location == "Czechoslovakia")] <- "Europe"
data$continent[which(is.na(data$continent))] <- "Asia" # 3 different Yemens
data$continent <- as.factor(data$continent)

# renaming the columns
names <- c("campaign","population_involved","success","max_participants","ID","year_max_involvement","first_year","last_year","location","pop_during_height", "more_than_3.5_involved","violent","continent")
colnames(data) <- names

# creating a "duration" variable instead of 3 variables representing years
data$duration <- data$last_year-data$first_year +1
```


The last step I took, was creating a variable `duration` which stores the length of each campaign in years.

To conclude this first step in processing and transformation I renamed the column of the data set to be easier to use.

Next we look at some variables which might be a good choice for a log transformation, namely `population_involved`, `max_participants` and `pop_during_height`. As can be seen there is a very large skew in them so a transformation makes sense.
```{r}
#| echo: false
par(mfrow=c(3,1))

hist(data$population_involved)
hist(data$max_participants)
hist(data$pop_during_height)
```
After the transformation the data looks the following. There is only one problem: when transforming the population involved we will get "-Inf" values as there are some which constitute less than 0.01% of the population, thus being 0, and the log2(0) not being defined. I try to create a dummy variable `pop_inv_zero` which is 1 if the value is "-Inf" and change the value in the log transformed `population_involved` to 0. This should, if my memory serves me correctly, follow the example given in class to deal with this behavior.

```{r}
#| echo: false
par(mfrow=c(3,1))


hist(log2(data$population_involved))
hist(log2(data$max_participants))
hist(log2(data$pop_during_height))

data$population_involved <- log2(data$population_involved)
data$max_participants <- log2(data$max_participants)
data$pop_during_height <- log2(data$year_max_involvement)
data$pop_inv_zero <- ifelse(data$population_involved == "-Inf",1,0)
data$population_involved[which(data$population_involved == "-Inf")] <- 0
```


# Fitting a model based on submodel-test, Anova and residual plots

I choose `success`as my response variable and `violent` as the main covariate I want to investigate. I drop first of all the campaign name, as it has 299 levels and is not useful in finding an overall pattern, same with the ID column. Additional columns i drop are the start and end years of the campaign as well as the location, I keep the year when it was at it's height as some periods tend to coincide with more successful campaigns, as discussed in the beginning.

The model summary for the full model with interactions on violent is given as follows:
```{r}
#| echo: false
#| message: false
#| warning: false
data_mod <- data[,c(2,3,4,6,10,12,13,14,15)]


logit_full <- glm(success~violent*.,data = data_mod, family = binomial)
summary(logit_full)
```

Interactions do not seem to have any significance whatsoever for our analysis, which is supported by the result of the model-submodel test.
Using a Type 2 anova table I see that our Target variable `violent` is very significant, as are `duration` and `population_involved`.
In the next step I remove `continent` as it shows a p-value of 0.777.
```{r}
#| echo: false
#Anova(logit_full)
m1 <- glm(success~violent+., data = data_mod,family = binomial)
anova(logit_full,m1,test="LRT")

Anova(m1,type=2)
```
```{r}
#| echo: false
m2 <- glm(success~violent+.-continent,data = data_mod,family = binomial)

final <- glm(success~violent+.-continent-max_participants,data=data_mod,family=binomial)
anova(m2,final,test="LRT")
Anova(final)
summary(final)
```
Before dropping any more variables I want to look at the residual plot to make sure we are not kicking a variable which only needs a transformation to be significant for our analysis.
Residuals of all variables seem to fit very well, especially `pop_during_height` with only a small dip and the upper end of the scale, same with `year_max_involvement`. `population_involved` shows a very slight upwards trend, and `duration` a little dip around 5 and a tendency to decrease for values larger 25 years. 
`max_participants` gets it's downwards dip at the upper end from a single observation which is why I will not try to fit another distribution to it.
Due to the anova table suggesting it, and the summary showing no significance, I decided to drop `max_participants` as well for our final model, to make it even simpler without loosing much explainability.

```{r}
#| warning: false
#| message: false
#| error: false

data_mod2 <- data_mod
data_mod2$residuals <- rstandard(final)
data_mod2$continent <- as.numeric(data_mod$continent)

data_long <- data_mod2 %>%
  pivot_longer(cols = -residuals, names_to = "variable", values_to = "value")

ggplot(data_long, aes(x = value, y = residuals)) + 
  geom_point() + 
  geom_smooth() + 
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(x = "Name", y = "Value", title = "Multiple Plots for Each Variable")


```
For the residuals we can see that they don't fit the QQ-plot perfectly with the main deviation being in the lower part and around 0. As expected, the residuals form two groups in teh histogram, although they seem to differ quite a bit.

```{r}
#| warning: false
#| echo: false
hist(data_mod2$residuals,freq=F,bins=20)
plot(final,which = c(2))
```
```{r}
#| include: false
#could not get the simulate() to return the same length of data as my data has

#B <- 1000
#Devs <- rep(0,B)
#for(b in 1:B){
#  newData <- data_mod
#  newData$success <- simulate(final)
#  newmodel <- glm(success~violent+.-continent-max_participants, data=newData, family=binomial)
#  Devs[b] <- newmodel$deviance
#}
#hist(Devs,main="", xlab="deviance")
#abline(v=final$deviance, col="red", lwd=2)
```


Surprisingly, when calling summary on the model, we see that both `violent` and `pop_inv_zero` are not at all significant, but the Anova table tells us that they are very significant and we can't drop them.

```{r}
#| echo: false
summary(final)
```

# Comparison to other models

To check whether the model really holds up I will compare it to a stepwise model based on the AIC, and a fitted model based on the duration of the campaign instead of the violence, to see if the focus on this variable is holding back our predictive ability.
In the plots below you can see again the residuals against the variables for the model on `duration`. The plots are very similar to the ones seen before and I do not see a fitting transformation to improve the fit of it.

```{r}
#| output: false
stepwise <- step(logit_full,direction = "backward")
summary(stepwise)
```

```{r}
#| echo: false
#| include: false
dm1 <- glm(success~duration*.,data=data_mod,family=binomial)
Anova(dm1)
dm2 <- glm(success~duration*population_involved+.,data=data_mod,family=binomial)
Anova(dm2)

dm3 <- glm(success~duration*population_involved+.-continent-max_participants,data=data_mod,family=binomial)

data_mod2 <- data_mod
data_mod2$residuals <- rstandard(dm3)
data_mod2$continent <- as.numeric(data_mod$continent)

data_long <- data_mod2 %>%
  pivot_longer(cols = -residuals, names_to = "variable", values_to = "value")
```
```{r}
#| warning: false
ggplot(data_long, aes(x = value, y = residuals)) + 
  geom_point() + 
  geom_smooth() + 
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(x = "Name", y = "Value", title = "Multiple Plots for Each Variable")


summary(dm3)
Anova(dm3)
```

For the first comparison I use the AICc and find that the model based on `duration` outperforms the final model and the stepwise model slightly, the two of them being identical. The full model comes in last by a large margin.
```{r}
AICc(final)
AICc(dm3)
AICc(stepwise)
AICc(logit_full)
```
I decided to use 5-Fold cross validation for the next comparison. Since I have 299 observations and a even split is not possible, I reuse one random observation form the dataset to fill it to 300. 
After calculating the average MSE of the models we see that the final model we built actually comes out on top, followed by the duration model, the stepwise model, and lastly the full model.

```{r}
#| inlcude: false
#| message: false
#| warning: false
#| output: false
# 5 fold CV
set.seed(1602)
K <- 5
MSE <- array(0,c(4,K))
fold <- matrix(c(sample(1:299),sample(1:299,1)),nrow=K) # because 299 obs i recycle a single one for CV

#success~duration*population_involved+.-continent-max_participants

for(k in 1:K){
  train <- data_mod[-fold[k,],]
  test <- data_mod[fold[k,],]
  final <- glm(success~violent+.-continent-max_participants,data=train,family=binomial)
  dm3 <- glm(success~duration*population_involved+.-continent-max_participants,data=train,family=binomial)
  logit_full <- glm(success~violent*.,data = train, family = binomial)
  stepwise <- step(logit_full,direction = "backward")
  pred_f <- predict(final,newdata=test,type="response")
  pred_d <- predict(dm3,newdata=test,type="response")
  pred_s <- predict(stepwise,newdata=test,type="response")
  pred_a <- predict(logit_full,newdata = test,type="response")
  MSE[1,k] <- sum((test$success-pred_f)^2)
  MSE[2,k] <- sum((test$success-pred_d)^2)
  MSE[3,k] <- sum((test$success-pred_s)^2)
  MSE[4,k] <- sum((test$success-pred_a)^2)
}
```

```{r}
rowMeans(MSE)
```

# Conclusion and Discussion

To sum up, the final model I decided on is $$ \mathbb{E}Y_n = \beta_0 + \beta_1 \mathbb{I}_{[violent]} + \beta_2 z_1 + \beta_3 x_1 + \beta_4 z_2 + \beta_5 x_2 + \beta_6 \mathbb{I}_{[pop\_inv\_zero]}  $$

Where we have Indicator functions for violence and if the population involved is below 0.01%, $x_1$ being the year of max involvement, $x_2$ the duration, $z_1$ and $z_2$ the log transformed population involved and population during the height of the campaign respectively.

Even though the Anova type 2 table suggests to keep all remaining variables and places high significance on them, it is hard to interpret the model as the remaining variables are mostly not significant in the summary statistic. 

All in all the fitted model does seem to perform the best out of the ones tested. What could be investigated further are interactions between variables, but from what has been seen in the anova tables and summary statistics they do not seem to hold much value for the analysis. Another aspect worth investigating could be starting from a dataset without any transformations and having a look, if we arrive at a different, maybe better model than the one we have now.

A larger dataset would also help improve the model as the current one only includes 299 campaigns, with an even split between successful and unsuccessful, but only 12 of them were violent, which makes it difficult to have as a significant variable.
Other than that, the most significant variable is `population_involved`, which does not seem very surprising, as the more people are campaigning the harder it is to put it down and not agree to their demands. Next up is duration, which has a negative estimate, meaning that the longer a campaign is going on for the more the log odds decrease. 
The `year_of_max_involvement` and `pop_during_height` are almost significant, the latter I guess could be due to some colinearity with the other population metrics. 





